---
title: "Diagnostics and tuning"
---

{{< include _setup.qmd >}}

You don't observe $\alpha_0$ on real data, so the validation curve from `early_stopping_rounds` is the only training-time signal. After fitting, `.diagnose()` reports magnitude statistics and warnings about extreme tail values, near-positivity violations, and single-point extrapolation. Those warnings often matter more for downstream variance than the bulk fit quality.

This page walks through `.diagnose()` on an under-tuned booster (so the warnings fire) and maps each warning to the hyperparameter that fixes it.

## Setup: a low-overlap dataset

We push the propensity close to 0 in part of the covariate space. This is where naive inverse-propensity weighting falls apart.

::: {.panel-tabset group="lang"}

## Python

```{python}
import numpy as np, pandas as pd
from rieszboost import RieszBooster, ATE

rng = np.random.default_rng(0)
n = 2000
x = rng.uniform(0, 1, n)
pi = 1 / (1 + np.exp(-(8.0 * x - 5.0)))
a = rng.binomial(1, pi)
df = pd.DataFrame({"a": a.astype(float), "x": x})
print(f"min π(x) = {pi.min():.3f}, fraction with π(x) < 0.05 = {(pi < 0.05).mean():.2f}")
```

## R

```{r}
set.seed(0)
n  <- 2000
x  <- runif(n)
pi <- 1 / (1 + exp(-(8 * x - 5)))
a  <- rbinom(n, 1, pi)
df <- data.frame(a = as.numeric(a), x = x)
cat(sprintf("min pi(x) = %.3f, fraction with pi(x) < 0.05 = %.2f\n",
            min(pi), mean(pi < 0.05)))
```

:::

## A greedy fit

Deep trees, no regularization, no early stopping. This will extrapolate badly in the low-π region.

::: {.panel-tabset group="lang"}

## Python

```{python}
greedy = RieszBooster(
    estimand=ATE(),
    n_estimators=300,
    learning_rate=0.3,
    max_depth=8,
    reg_lambda=0.0,
).fit(df)

print(greedy.diagnose(df).summary())
```

## R

```{r}
greedy <- RieszBooster$new(
  estimand = ATE(),
  n_estimators = 300L,
  learning_rate = 0.3,
  max_depth = 8L,
  reg_lambda = 0.0
)
greedy$fit(df)
cat(greedy$diagnose(df)$summary)
```

:::

## Reading the report

The summary has three parts.

**Magnitude block.** RMS, mean, min/max, and `|alpha|` quantiles. The 99th percentile and the max are the load-bearing numbers. Healthy fits have max no more than a few times the 99th percentile. Max far above the 99th percentile signals a single-point extrapolation outlier: one leaf is far from any training data and its prediction has run away.

**Extreme rows.** Rows where $|\hat\alpha|$ exceeds `extreme_threshold` (default 30). Downstream IF estimator variance is dominated by the largest few weights, so even 1% of rows over the threshold inflates confidence intervals.

**Warnings.** Two automatic checks:

1. *"X% of rows have $|\hat\alpha| >$ threshold — possible near-positivity violation."* Either the data has genuine near-positivity (truncate `extreme_threshold` and report the loss of overlap), or the booster is over-fitting (tune as below).
2. *"max $|\hat\alpha|$ is >10× the 99th percentile — likely a single extrapolation outlier."* The booster found a leaf with no nearby training data. Tighten trees, increase `reg_lambda`, or stop earlier.

## A tuned fit

Same data with sensible hyperparameters and early stopping.

::: {.panel-tabset group="lang"}

## Python

```{python}
tuned = RieszBooster(
    estimand=ATE(),
    n_estimators=2000,
    learning_rate=0.05,
    max_depth=3,
    reg_lambda=1.0,
    early_stopping_rounds=20,
    validation_fraction=0.2,
).fit(df)

print(f"stopped at iter {tuned.best_iteration_}, "
      f"best held-out Riesz loss = {tuned.best_score_:.4f}")
print()
print(tuned.diagnose(df).summary())
```

## R

```{r}
tuned <- RieszBooster$new(
  estimand = ATE(),
  n_estimators = 2000L,
  learning_rate = 0.05,
  max_depth = 3L,
  reg_lambda = 1.0,
  early_stopping_rounds = 20L,
  validation_fraction = 0.2
)
tuned$fit(df)
cat(sprintf("stopped at iter %d, best held-out Riesz loss = %.4f\n\n",
            reticulate::py_to_r(tuned$py$best_iteration_),
            reticulate::py_to_r(tuned$py$best_score_)))
cat(tuned$diagnose(df)$summary)
```

:::

The max $|\hat\alpha|$ shrinks by an order of magnitude, the extreme-row count drops, and the warnings clear.

## Tuning recipe

| Knob | Default | When to change | Effect |
|---|---|---|---|
| `max_depth` | 4 | Drop to 2–3 for extrapolation outliers | Shallow trees can't carve out a high-magnitude leaf for a single point. |
| `learning_rate` | 0.05 | Drop to 0.02 for harder problems | Smaller updates per tree; pair with higher `n_estimators`. |
| `n_estimators` + `early_stopping_rounds` | 200, `None` | Set `n_estimators=1000-2000` and `early_stopping_rounds=20` | The validation split picks the iteration count. |
| `validation_fraction` | 0.0 | Set to 0.2 with `early_stopping_rounds` | Internal split for early stopping. Alternative: pass `eval_set=` explicitly. |
| `reg_lambda` | 1.0 | Bump to 5–10 for low-overlap data | xgboost L2 on leaf weights; damps the magnitude of any single leaf. |
| `subsample` | 1.0 | Try 0.5–0.8 with very large $n$ | Stochastic boosting. |

`XGBoostBackend` exposes two more knobs:

| Knob | Default | When to change |
|---|---|---|
| `gradient_only` | `False` | Set `True` to reproduce Lee-Schuler Algorithm 2 / Friedman 2001 exactly. |
| `hessian_floor` | `2.0` | Lower bound on the per-row Hessian. The default matches the natural Hessian of original-data rows. |

See [backends and losses](backends-and-losses.qmd) for the backend protocol.

## What `.diagnose()` does not do

It does not estimate $\alpha_0$ for comparison; that closed form is generally unavailable. It does not assess sample size. It checks magnitude and the two failure modes above. For fit quality, track the held-out Riesz loss (`best_score_`). For downstream confidence intervals, use the bootstrap on the plug-in estimator.
