---
title: "Quickstart — ATE in 30 lines"
---

{{< include _setup.qmd >}}

This page fits a Riesz representer for the average treatment effect on a synthetic dataset. Pick the language tab on any code block; your choice persists across the site.

## Simulate

Binary treatment with known propensity $\pi(x)$. We will recover $\alpha_0(a, x) = a/\pi(x) - (1-a)/(1-\pi(x))$ without estimating $\pi$.

::: {.panel-tabset group="lang"}

## Python

```{python}
import numpy as np
import pandas as pd
from rieszboost import RieszBooster, ATE

rng = np.random.default_rng(0)
n = 4000
x = rng.uniform(0, 1, n)
pi = 1 / (1 + np.exp(-(-0.02 * x - x**2 + 4 * np.log(x + 0.3) + 1.5)))
a = rng.binomial(1, pi)
df = pd.DataFrame({"a": a.astype(float), "x": x})
df.head()
```

## R

```{r}
set.seed(0)
n  <- 4000
x  <- runif(n)
pi <- 1 / (1 + exp(-(-0.02 * x - x^2 + 4 * log(x + 0.3) + 1.5)))
a  <- rbinom(n, 1, pi)
df <- data.frame(a = as.numeric(a), x = x)
head(df)
```

:::

## Fit

Configure the booster on the constructor, then call `.fit(df)`. Early stopping uses an internal validation split sized by `validation_fraction`.

::: {.panel-tabset group="lang"}

## Python

```{python}
booster = RieszBooster(
    estimand=ATE(treatment="a", covariates=("x",)),
    n_estimators=2000,
    learning_rate=0.05,
    max_depth=4,
    early_stopping_rounds=20,
    validation_fraction=0.2,
)
booster.fit(df)
print(f"stopped at iteration {booster.best_iteration_}, "
      f"best held-out Riesz loss = {booster.best_score_:.4f}")
```

## R

```{r}
booster <- RieszBooster$new(
  estimand = ATE(treatment = "a", covariates = "x"),
  n_estimators = 2000L,
  learning_rate = 0.05,
  max_depth = 4L,
  early_stopping_rounds = 20L,
  validation_fraction = 0.2
)
booster$fit(df)
cat(sprintf(
  "stopped at iteration %d, best held-out Riesz loss = %.4f\n",
  reticulate::py_to_r(booster$py$best_iteration_),
  reticulate::py_to_r(booster$py$best_score_)
))
```

:::

## Predict and compare to the closed form

We know $\alpha_0$ here, so we can score the fit directly.

::: {.panel-tabset group="lang"}

## Python

```{python}
alpha_hat  = booster.predict(df)
true_alpha = a / pi - (1 - a) / (1 - pi)
print(f"corr(α̂, α₀) = {np.corrcoef(alpha_hat, true_alpha)[0, 1]:.3f}")
print(f"RMSE         = {np.sqrt(np.mean((alpha_hat - true_alpha)**2)):.3f}")
```

## R

```{r}
alpha_hat  <- booster$predict(df)
true_alpha <- a / pi - (1 - a) / (1 - pi)
cat(sprintf("corr(alpha_hat, alpha_0) = %.3f\n",
            cor(alpha_hat, true_alpha)))
cat(sprintf("RMSE                     = %.3f\n",
            sqrt(mean((alpha_hat - true_alpha)^2))))
```

:::

## Visual sanity check

::: {.panel-tabset group="lang"}

## Python

```{python}
#| fig-cap: "Estimated Riesz representer vs. the closed-form ground truth on the synthetic ATE DGP."
import matplotlib.pyplot as plt
fig, ax = plt.subplots()
ax.scatter(true_alpha, alpha_hat, s=4, alpha=0.4)
lim = max(abs(true_alpha).max(), abs(alpha_hat).max())
ax.plot([-lim, lim], [-lim, lim], "k--", lw=1)
ax.set_xlabel(r"$\alpha_0$ (closed-form)")
ax.set_ylabel(r"$\hat\alpha$ (rieszboost)")
plt.show()
```

## R

```{r}
#| fig-cap: "Estimated Riesz representer vs. the closed-form ground truth (R)."
lim <- max(abs(c(true_alpha, alpha_hat)))
plot(true_alpha, alpha_hat, pch = 16, cex = 0.4, col = adjustcolor("black", 0.4),
     xlim = c(-lim, lim), ylim = c(-lim, lim),
     xlab = expression(alpha[0]~"(closed-form)"),
     ylab = expression(hat(alpha)~"(rieszboost)"))
abline(0, 1, lty = 2)
```

:::

## Diagnostics

Inspect the fit before using $\hat\alpha$ downstream. `.diagnose()` reports the magnitude distribution, positivity warnings, and the held-out Riesz loss.

::: {.panel-tabset group="lang"}

## Python

```{python}
print(booster.diagnose(df).summary())
```

## R

```{r}
cat(booster$diagnose(df)$summary)
```

:::

## Next

- [scikit-learn integration](sklearn-integration.qmd) for out-of-fold predictions used in TMLE / one-step / DML.
- [Estimands](estimands.qmd) for the other built-ins and custom $m$.
- [Diagnostics and tuning](diagnostics-and-tuning.qmd) for reading `.diagnose()` and choosing hyperparameters.
