---
title: "scikit-learn integration"
---

{{< include _setup.qmd >}}

`RieszBooster` is an `sklearn.base.BaseEstimator`. It composes with `cross_val_predict`, `GridSearchCV`, `Pipeline`, `clone`, and the rest of `sklearn.model_selection`.

## Cross-fitting

Plug-in $\hat\alpha$ in a one-step / TMLE / DML estimator must be out-of-fold. Use `cross_val_predict`:

::: {.panel-tabset group="lang"}

## Python

```{python}
import numpy as np, pandas as pd
from sklearn.model_selection import cross_val_predict
from rieszboost import RieszBooster, ATE

rng = np.random.default_rng(0)
n = 2000
x = rng.uniform(0, 1, n)
pi = 1 / (1 + np.exp(-(-0.02 * x - x**2 + 4 * np.log(x + 0.3) + 1.5)))
a = rng.binomial(1, pi)
df = pd.DataFrame({"a": a.astype(float), "x": x})

booster = RieszBooster(
    estimand=ATE(),
    n_estimators=2000,
    early_stopping_rounds=20,
    validation_fraction=0.2,
    learning_rate=0.05,
    max_depth=4,
)

alpha_oof = cross_val_predict(booster, df, cv=5)
alpha_oof.shape, alpha_oof[:5]
```

## R

```{r}
set.seed(0)
n  <- 2000
x  <- runif(n)
pi <- 1 / (1 + exp(-(-0.02 * x - x^2 + 4 * log(x + 0.3) + 1.5)))
a  <- rbinom(n, 1, pi)
df <- data.frame(a = as.numeric(a), x = x)

# Call sklearn directly via reticulate, passing the underlying Python
# estimator from `booster$py`.
sk_ms <- reticulate::import("sklearn.model_selection", convert = FALSE)
booster <- RieszBooster$new(
  estimand = ATE(),
  n_estimators = 2000L,
  early_stopping_rounds = 20L,
  validation_fraction = 0.2,
  learning_rate = 0.05,
  max_depth = 4L
)
alpha_oof <- as.numeric(reticulate::py_to_r(
  sk_ms$cross_val_predict(booster$py, reticulate::r_to_py(df), cv = 5L)
))
length(alpha_oof); head(alpha_oof)
```

:::

Each fold runs its own internal early-stopping split.

## Hyperparameter search

`RieszBooster.score(X)` returns negative held-out Riesz loss, so `GridSearchCV` selects the configuration with the lowest loss.

::: {.panel-tabset group="lang"}

## Python

```{python}
from sklearn.model_selection import GridSearchCV

grid = GridSearchCV(
    RieszBooster(
        estimand=ATE(),
        n_estimators=300,
        early_stopping_rounds=20,
        validation_fraction=0.2,
    ),
    param_grid={"learning_rate": [0.03, 0.05, 0.1], "max_depth": [3, 4, 5]},
    cv=3,
    n_jobs=1,
).fit(df)

print("best params:", grid.best_params_)
print(f"best score (neg Riesz loss): {grid.best_score_:.4f}")

best = grid.best_estimator_
alpha_hat = best.predict(df)
alpha_hat[:5]
```

## R

```{r}
sk_ms <- reticulate::import("sklearn.model_selection", convert = FALSE)
base  <- RieszBooster$new(
  estimand = ATE(),
  n_estimators = 300L,
  early_stopping_rounds = 20L,
  validation_fraction = 0.2
)
grid <- sk_ms$GridSearchCV(
  base$py,
  param_grid = reticulate::dict(
    learning_rate = c(0.03, 0.05, 0.1),
    max_depth = list(3L, 4L, 5L)
  ),
  cv = 3L,
  n_jobs = 1L
)
grid$fit(reticulate::r_to_py(df))
cat("best params:\n"); print(reticulate::py_to_r(grid$best_params_))
cat(sprintf("best score (neg Riesz loss): %.4f\n",
            reticulate::py_to_r(grid$best_score_)))
```

:::

## Other sklearn integrations

`RieszBooster` exposes the standard `BaseEstimator` interface:

- `clone(booster)` returns a fresh unfitted copy with the same hyperparameters.
- `Pipeline` wraps a feature-engineering step in front of the booster.
- `HalvingGridSearchCV`, `RandomizedSearchCV`, `BayesSearchCV` (skopt) work without modification.
- `get_params()` / `set_params()` are used by the search classes to introspect and mutate the configuration.

## See also

- [`examples/lee_schuler/binary_dgp.py`](https://github.com/rieszreg/rieszboost/blob/main/examples/lee_schuler/binary_dgp.py) builds an EEE estimator using `cross_val_predict` to produce out-of-fold $\hat\alpha$.
- [Diagnostics and tuning](diagnostics-and-tuning.qmd) covers which hyperparameters to vary in a grid search.
