---
title: "Forest backend (forestriesz)"
---

{{< include ../_setup.qmd >}}

`forestriesz` ships two random-forest backends. They differ in whether the user has to make per-estimand modeling choices to fit the model.

- **`AugForestRieszRegressor`** — augmentation-style. An ensemble of single-tree Riesz regressors fit on the augmented dataset of evaluation points $z_r$ with weights $(D_r, C_r)$ that `Estimand.augment` produces. Each tree uses sklearn-standard hyperparameters and a loss-aware splitter that handles every built-in Bregman loss directly. **Works on every estimand without per-estimand configuration** — built-in estimands (`ATE`, `ATT`, `TSM`, `AdditiveShift`, `LocalShift`) and any user-defined `Estimand` use the same call.
- **`ForestRieszRegressor`** — moment-style. Implements [Chernozhukov, Newey, Quintas-Martínez, Syrgkanis (ICML 2022)](https://proceedings.mlr.press/v162/chernozhukov22a/chernozhukov22a.pdf). To fit the model the user must supply a list of basis functions of the data — the `riesz_feature_fns` parameter — that the forest combines linearly inside each leaf. The required list is per-estimand: `ATE`, `ATT`, and `TSM` get reasonable defaults from `riesz_feature_fns="auto"`, but other estimands need a user-constructed list. Supports honest-split confidence intervals via `predict_interval` for single-basis fits.

`rieszreg` exposes the two entry points as `Backend.fit_augmented` and `MomentBackend.fit_rows` respectively.

## How they work

### Augmentation-style (`AugForestRieszRegressor`)

For each original row $i$, `Estimand.augment` emits the original observation $Z_i$ and one evaluation point per linear-form term in $m(\mu)(Z_i, Y_i)$. Each augmented row $r$ carries an evaluation point $z_r$, a weight $D_r \in \{0, 1\}$ indicating whether $z_r$ is the original observation, and a coefficient $C_r$ collecting the trace coefficients at $z_r$. The empirical Bregman-Riesz loss is

$$
\hat L(\alpha) \;=\; \frac{1}{n} \sum_{r=1}^M \big[\, D_r\, \tilde h(\alpha(z_r)) \;+\; C_r\, h'(\alpha(z_r)) \,\big].
$$

See [Estimands](../estimands.qmd) for the construction. Each tree partitions the feature space and stores the loss-optimal $\alpha$ in every leaf $\ell$:

$$
\hat\alpha_\ell \;=\; \arg\min_{\alpha} \sum_{r \in \ell} \big[\, D_r\, \tilde h(\alpha) \;+\; C_r\, h'(\alpha) \,\big].
$$

For `SquaredLoss` this has the closed form $\hat\alpha_\ell = -\sum_{r \in \ell} C_r / \sum_{r \in \ell} D_r$; for `KLLoss`, `BernoulliLoss`, and `BoundedSquaredLoss` the per-leaf optimum has its own closed form (see [Tree backend](tree.qmd) for the derivations). The forest averages the per-tree predictions of $\hat\alpha$.

Because $D_r$ and $C_r$ already vary across augmented rows for every estimand — original observations have $D_r = 1, C_r = 0$ and counterfactual evaluation points have $D_r = 0, C_r \ne 0$ — the splitter learns from the full feature space directly. The user does not supply any basis functions.

### Moment-style (`ForestRieszRegressor`)

The user supplies a list of functions $\varphi = (\varphi_1, \dots, \varphi_p)$ of the data. Inside each leaf $\ell$ the per-leaf parameter $\theta_\ell \in \mathbb R^p$ solves the linear-moment equation

$$
\hat\theta_\ell \;=\; \big(\textstyle\sum_{i\in\ell} \varphi(Z_i)\,\varphi(Z_i)^{\top}\big)^{-1}
                 \textstyle\sum_{i\in\ell} m(\varphi)(Z_i, Y_i),
$$

and predictions are $\hat\alpha(z) = \hat\theta(z_{\text{split}}) \cdot \varphi(z)$. The forest is fit on the $n$ original rows, and the MSE splitting criterion picks splits to minimize the sum of in-leaf residuals against this leaf optimum.

The list $\varphi$ has to be chosen per estimand. The default `riesz_feature_fns="auto"` picks defaults for built-in estimands:

| Estimand | Default list | Effect |
|---|---|---|
| `ATE`, `ATT` | `[1{A=0}, 1{A=1}]` | Forest splits on covariates only; the basis resolves treatment in each leaf. |
| `TSM(level=v)` | `[1{A=v}]` | Single-basis fit; intervals supported. |
| `AdditiveShift`, `LocalShift`, custom | none — auto-resolution falls back to a constant function, which is row-degenerate | Backend raises with a hint. Use `AugForestRieszRegressor` instead, or pass a custom list. |

For ATE the leaf solve recovers $1/\hat P(A = a \mid X\text{-leaf})$ — the IPW representer.

## How they differ

| | `AugForestRieszRegressor` | `ForestRieszRegressor` |
|---|---|---|
| Per-estimand setup | none — works on every estimand directly | user supplies `riesz_feature_fns` (auto-resolved for `ATE`, `ATT`, `TSM`) |
| Loss support | `SquaredLoss`, `KLLoss`, `BernoulliLoss`, `BoundedSquaredLoss` | `SquaredLoss` only |
| Confidence intervals | not in v1 | yes — `predict_interval` on single-basis fits with `honest=True, inference=True` |
| Splits on | full feature space | covariates only when the basis handles treatment |
| Backend Protocol | `Backend.fit_augmented` | `MomentBackend.fit_rows` |
| When to pick | default for general use; required for shift-style or custom estimands | when you need CIs on `ATE`, `ATT`, or `TSM` |

## Bregman losses

`AugForestRieszRegressor` accepts every loss `rieszreg` ships. The per-tree splitter dispatches to a per-loss leaf-loss formula (squared, KL, Bernoulli, bounded-squared) so splits are chosen against the loss the user actually supplied. The forest averages per-tree $\hat\alpha$ predictions.

```{python}
import numpy as np, pandas as pd
from forestriesz import AugForestRieszRegressor, TSM
from rieszreg import KLLoss

rng = np.random.default_rng(0)
n = 1500
x = rng.uniform(0, 1, n)
pi = 1 / (1 + np.exp(-(0.5 * x - 0.3)))
a = rng.binomial(1, pi).astype(float)
df_tsm = pd.DataFrame({"a": a, "x": x})

# KLLoss is matched to density-ratio estimands like TSM and IPSI;
# it requires non-negative C-coefficients in the augmented dataset.
est = AugForestRieszRegressor(
    estimand=TSM(level=1, treatment="a", covariates=("x",)),
    loss=KLLoss(),
    n_estimators=200,
    min_samples_leaf=10,
    random_state=0,
)
est.fit(df_tsm)
alpha_hat = est.predict(df_tsm)
print(f"alpha_hat range: [{alpha_hat.min():.3f}, {alpha_hat.max():.3f}]")
```

`KLLoss` and `BernoulliLoss` reject `ATE` / `ATT` / shift-style data at fit time — they require non-negative $C$-coefficients, which difference-style estimands violate. `BoundedSquaredLoss(lo, hi)` accepts any signed coefficients and clips $\hat\alpha$ into the interval.

## Quickstart — ATE

::: {.panel-tabset group="lang"}

## Python

```{python}
import numpy as np, pandas as pd
from forestriesz import ForestRieszRegressor, ATE

rng = np.random.default_rng(0)
n = 1500
x = rng.uniform(0, 1, n)
pi = 1 / (1 + np.exp(-(-0.02*x - x**2 + 4*np.log(x + 0.3) + 1.5)))
a = rng.binomial(1, pi).astype(float)
df = pd.DataFrame({"a": a, "x": x})

fr = ForestRieszRegressor(
    estimand=ATE(treatment="a", covariates=("x",)),
    n_estimators=500,
    min_samples_leaf=10,
    random_state=0,
)
fr.fit(df)

alpha_hat  = fr.predict(df)
true_alpha = a / pi - (1 - a) / (1 - pi)
print(f"corr(α̂, α₀) = {np.corrcoef(alpha_hat, true_alpha)[0, 1]:.3f}")
print(f"RMSE         = {np.sqrt(np.mean((alpha_hat - true_alpha)**2)):.3f}")
```

## R

```{r}
#| eval: false
pkgload::load_all("../packages/forestriesz/r/forestriesz")
use_python_forestriesz(file.path(getwd(), "../.venv/bin/python"))

set.seed(0)
n  <- 1500
x  <- runif(n)
pi <- 1 / (1 + exp(-(-0.02 * x - x^2 + 4 * log(x + 0.3) + 1.5)))
a  <- rbinom(n, 1, pi)
df <- data.frame(a = as.numeric(a), x = x)

# ATE / ATT use a multi-basis list, which is Python-only in v1.
# Call into Python directly via reticulate.
fr_py <- reticulate::import("forestriesz", convert = FALSE)
pd_py <- reticulate::import("pandas", convert = FALSE)
py_df <- pd_py$DataFrame(reticulate::r_to_py(list(a = as.numeric(a), x = x)))

fr <- fr_py$ForestRieszRegressor(
  estimand = fr_py$ATE(treatment = "a", covariates = list("x")),
  n_estimators = 500L, min_samples_leaf = 10L, random_state = 0L
)
fr$fit(py_df)
alpha_hat <- as.numeric(reticulate::py_to_r(fr$predict(py_df)))
```

:::

## Quickstart — `AdditiveShift` with the augmentation-style backend

`AdditiveShift` has no canonical list of basis functions, so the moment-style backend would require the user to construct one. `AugForestRieszRegressor` works on it directly.

```{python}
from forestriesz import AugForestRieszRegressor, AdditiveShift

# Continuous-treatment data
a_cont = rng.normal(0.5*x, 1.0)
df_cont = pd.DataFrame({"a": a_cont, "x": x})

afr = AugForestRieszRegressor(
    estimand=AdditiveShift(delta=0.5, treatment="a", covariates=("x",)),
    n_estimators=500,
    min_samples_leaf=10,
    random_state=0,
)
afr.fit(df_cont)

alpha_hat = afr.predict(df_cont)
print(f"alpha_hat range: [{alpha_hat.min():.3f}, {alpha_hat.max():.3f}]")
```

## Quickstart — TSM with confidence intervals

`TSM` uses a single-basis fit, so v1's `predict_interval` works.

::: {.panel-tabset group="lang"}

## Python

```{python}
from forestriesz import ForestRieszRegressor, TSM

fr = ForestRieszRegressor(
    estimand=TSM(level=1, treatment="a", covariates=("x",)),
    n_estimators=500,
    honest=True,
    inference=True,
    random_state=0,
)
fr.fit(df)

alpha_hat = fr.predict(df)
lb, ub    = fr.predict_interval(df, alpha=0.05)
print(f"mean 95% CI width = {(ub - lb).mean():.3f}")
print(f"mean coverage of α̂ in [lb, ub] = {((lb <= alpha_hat) & (alpha_hat <= ub)).mean():.2f}")
```

## R

```{r}
#| eval: false
pkgload::load_all("../packages/forestriesz/r/forestriesz")
use_python_forestriesz(file.path(getwd(), "../.venv/bin/python"))

fr <- ForestRieszRegressor$new(
  estimand = TSM(level = 1L, treatment = "a", covariates = "x"),
  n_estimators = 500L,
  honest = TRUE,
  inference = TRUE,
  random_state = 0L
)
fr$fit(df)
alpha_hat <- fr$predict(df)
ci <- fr$predict_interval(df, alpha = 0.05)
```

:::

## Hyperparameters

### `AugForestRieszRegressor`

Mirrors [`sklearn.ensemble.RandomForestRegressor`](https://scikit-learn.org/stable/modules/generated/sklearn.ensemble.RandomForestRegressor.html):

| Knob | Default | Notes |
|---|---|---|
| `n_estimators` | 100 | Number of trees. |
| `max_depth` | None | Grow until leaves saturate `min_samples_leaf`. |
| `min_samples_split` | 2 | Minimum count of original ($D > 0$) augmented rows in a node before considering a split. |
| `min_samples_leaf` | 1 | Minimum count of original rows in each child of a candidate split. |
| `max_features` | 1.0 | Per-split feature-subsampling rule. |
| `max_leaf_nodes` | None | Cap on per-tree leaf count. |
| `min_impurity_decrease` | 0.0 | Reject splits below this gain. |
| `ccp_alpha` | 0.0 | Cost-complexity pruning per tree. |
| `bootstrap` | True | Per-tree resampling at the original-row level (block bootstrap). |
| `max_samples` | None | Per-tree subsample size as fraction or count of original rows. |
| `n_jobs` | None | Parallelism across trees via `joblib.Parallel`. |
| `splitter` | `"exact"` | `"exact"`, `"hist"`, `"random"`, or `"python"` — riesztree splitter dispatch. |
| `max_bins` | 255 | Histogram-splitter bin count. |
| `categorical_features` | None | Column indices treated as integer category labels. |

### `ForestRieszRegressor`

Forwarded to EconML's `BaseGRF`:

| Knob | Default | Notes |
|---|---|---|
| `n_estimators` | 100 | Number of trees. With `inference=True` must be divisible by `subforest_size` (default 4). |
| `max_depth` | None | Grow until leaves saturate `min_samples_leaf`. |
| `min_samples_leaf` | 5 | Tighter leaves give lower bias, higher variance. Paper guidance: 5–20. |
| `min_samples_split` | 10 | |
| `max_samples` | 0.45 | Per-tree subsample fraction. GRF default. |
| `honest` | False | Enable for valid-coverage CIs (50/50 sample split). |
| `inference` | False | Enable for `predict_interval`; preserves subforest structure. |
| `subforest_size` | 4 | Trees per subforest when computing inference variance. |
| `l2` | 0.01 | Ridge added to the per-leaf Jacobian. |
| `n_jobs` | -1 | Parallel tree training. |

## Save / load

```{python}
import tempfile, os
with tempfile.TemporaryDirectory() as tmp:
    save_dir = os.path.join(tmp, "fitted")
    fr.save(save_dir)

    loaded = ForestRieszRegressor.load(save_dir)
    pred_loaded = loaded.predict(df)
    print(f"max abs diff after round-trip: {np.max(np.abs(pred_loaded - alpha_hat)):.2e}")
```

For built-in estimands the moment-style basis is auto-resolved on load. For custom bases, repass the callables: `ForestRieszRegressor.load(path, riesz_feature_fns=my_basis)`. `AugForestRieszRegressor` round-trips without any extra arguments.

## Sklearn integration

Both regressors are `BaseEstimator`s, so they compose with `clone`, `GridSearchCV`, `cross_val_predict`, `Pipeline`. See [Tuning and Cross-Fitting](../tuning-and-cross-fitting.qmd).

## When the constant basis raises (moment-style only)

For built-in estimands the per-row moment $m(\varphi)(Z_i, Y_i)$ doesn't depend on $Z_i$ when $\varphi \equiv 1$, so the constant basis is degenerate (every leaf would predict the same constant). `ForestRieszRegressor` detects this and raises with a hint to use `riesz_feature_fns="auto"` (the default) or supply a custom list. Explicit `riesz_feature_fns=None` is what raises. `AugForestRieszRegressor` is unaffected — its splitter does not depend on a user-supplied basis.
