---
title: "Forest backend (forestriesz)"
---

{{< include ../_setup.qmd >}}

`forestriesz` ships two backends, both built on EconML's `BaseGRF` (the same Cython criterion the paper authors used):

- **`ForestRieszRegressor`** — moment-style. Implements [Chernozhukov, Newey, Quintas-Martínez, Syrgkanis (ICML 2022)](https://proceedings.mlr.press/v162/chernozhukov22a/chernozhukov22a.pdf). Trains the GRF on the n original rows; needs a sieve (auto-resolved for ATE/ATT/TSM). Supports `predict_interval` for single-basis fits.
- **`AugForestRieszRegressor`** — augmentation-style. Trains the GRF on the M = k·n augmented `(a, b)` evaluation points produced by `Estimand.augment`. **Estimand-agnostic** — works on every built-in estimand and any custom user `Estimand` without a sieve. The combination of GRF + augmented data appears to be novel; see [How they differ](#how-they-differ) below.

`rieszreg` exposes the two entry points as `MomentBackend.fit_rows` and `Backend.fit_augmented` respectively.

## How they work

### Moment-style (`ForestRieszRegressor`)

In each leaf $\ell$ the per-leaf parameter $\theta_\ell$ solves the linear-moment equation

$$
\hat\theta_\ell = \big(\textstyle\sum_{i\in\ell} \varphi(W_i)\,\varphi(W_i)^{\top}\big)^{-1}
                 \textstyle\sum_{i\in\ell} m(W_i;\,\varphi),
$$

and predictions are $\hat\alpha(z) = \hat\theta(z_{\text{split}}) \cdot \varphi(z)$. The MSE splitting criterion picks splits to minimize the sum of in-leaf residuals against this leaf optimum.

The user-supplied basis $\varphi = (\varphi_1, \dots, \varphi_p)$ controls expressivity. The default `riesz_feature_fns="auto"` picks a sensible sieve for built-in estimands:

| Estimand | Default sieve | Effect |
|---|---|---|
| `ATE`, `ATT` | `[1{T=0}, 1{T=1}]` | Forest splits on covariates only; sieve resolves treatment in each leaf. |
| `TSM(level=v)` | `[1{T=v}]` | Single-basis sieve; intervals supported. |
| `AdditiveShift`, `LocalShift`, custom | none | Constant basis is row-degenerate; raises with a hint. Use `AugForestRieszRegressor` instead. |

### Augmentation-style (`AugForestRieszRegressor`)

The augmentation engine produces M = k·n evaluation points $z_k$ with coefficients $(a_k, b_k)$ such that the squared Riesz loss is $\sum_k a_k\,\alpha(z_k)^2 + b_k\,\alpha(z_k)$. Per *augmented* row:

$$
J_k = 2 a_k \cdot \varphi(z_k)\varphi(z_k)^{\top}, \qquad A_k = -b_k \cdot \varphi(z_k).
$$

Originals contribute $a_k = 1, b_k = 0$ so $J_k$ varies across rows; counterfactual evaluation points contribute $a_k = 0, b_k \neq 0$ so $A_k$ varies. The forest can split the full feature space usefully even with the constant basis $\varphi \equiv 1$. For ATE the leaf solve recovers $1/\hat P(T = t \mid X\text{-leaf})$; for AdditiveShift it recovers a local density-ratio estimator.

The trade-off is that the augmented training rows aren't iid — there are blocks of $k$ correlated rows per original $W_i$ — so GRF's variance estimator no longer applies and `predict_interval` is not exposed in v1. Cluster-robust variance with `origin_index` as the cluster id is on the v2 roadmap.

## How they differ

| | `ForestRieszRegressor` | `AugForestRieszRegressor` |
|---|---|---|
| Estimand support | ATE/ATT/TSM via auto sieve; others need user-supplied sieve | Every built-in + custom user `Estimand`, no sieve |
| Loss support | `SquaredLoss` only | All four built-in losses (`SquaredLoss`, `KLLoss`, `BernoulliLoss`, `BoundedSquaredLoss`) |
| `predict_interval` | yes (single-basis fits, `honest=True`) | not in v1 |
| Per-fit speed | baseline | ≈1.0× at n≤2000, ≈1.15× at n=5000 |
| RMSE on shared estimands (ATE, TSM) | baseline | within ~5% |
| Splits on | covariates only when sieve handles treatment | full (T, X) feature space |
| Backend Protocol | `MomentBackend.fit_rows` | `Backend.fit_augmented` |

Default to `ForestRieszRegressor` for ATE/ATT/TSM when you want CIs. Reach for `AugForestRieszRegressor` for shift-style estimands, custom moments, non-quadratic Bregman losses, or when you want one estimator that just works on everything. See [`test_aug_vs_moment.py`](https://github.com/rieszreg/forestriesz/blob/main/python/tests/test_aug_vs_moment.py) for the benchmark behind the speed and RMSE numbers.

## Bregman losses on the augmentation-style backend

`AugForestRieszRegressor` accepts every loss `rieszreg` ships. The fit still uses the squared MSE criterion to grow the trees (splits that maximize variance reduction in $-\Sigma b / (2 \Sigma a)$ also separate the monotonically-related Bregman optima well in practice), then a post-hoc Newton iteration on each leaf's augmented rows replaces the stored θ with the Bregman-optimal value. For locally constant fits the Newton has the closed form $\alpha^* = -B/(2A)$ followed by the loss's link; for sieves it's a small p×p Newton.

```{python}
import numpy as np, pandas as pd
from forestriesz import AugForestRieszRegressor, TSM
from rieszreg import KLLoss

rng = np.random.default_rng(0)
n = 1500
x = rng.uniform(0, 1, n)
pi = 1 / (1 + np.exp(-(0.5 * x - 0.3)))
a = rng.binomial(1, pi).astype(float)
df_tsm = pd.DataFrame({"a": a, "x": x})

# KLLoss enforces α̂ > 0 via the exp link — matched to density-ratio
# estimands like TSM and IPSI. Requires non-negative m-coefficients.
est = AugForestRieszRegressor(
    estimand=TSM(level=1, treatment="a", covariates=("x",)),
    loss=KLLoss(),
    n_estimators=200,
    min_samples_leaf=10,
    random_state=0,
)
est.fit(df_tsm)
alpha_hat = est.predict(df_tsm)
print(f"all positive: {bool((alpha_hat > 0).all())}")
print(f"min/max: {alpha_hat.min():.3f} / {alpha_hat.max():.3f}")
```

`KLLoss` and `BernoulliLoss` reject ATE / ATT / shift-style data at fit time (they require non-negative m-coefficients, which difference-of-evaluations estimands violate). `BoundedSquaredLoss(lo, hi)` accepts any signed coefficients and clips α̂ into the interval.

## Quickstart — ATE

::: {.panel-tabset group="lang"}

## Python

```{python}
import numpy as np, pandas as pd
from forestriesz import ForestRieszRegressor, ATE

rng = np.random.default_rng(0)
n = 1500
x = rng.uniform(0, 1, n)
pi = 1 / (1 + np.exp(-(-0.02*x - x**2 + 4*np.log(x + 0.3) + 1.5)))
a = rng.binomial(1, pi).astype(float)
df = pd.DataFrame({"a": a, "x": x})

fr = ForestRieszRegressor(
    estimand=ATE(treatment="a", covariates=("x",)),
    n_estimators=500,
    min_samples_leaf=10,
    random_state=0,
)
fr.fit(df)

alpha_hat  = fr.predict(df)
true_alpha = a / pi - (1 - a) / (1 - pi)
print(f"corr(α̂, α₀) = {np.corrcoef(alpha_hat, true_alpha)[0, 1]:.3f}")
print(f"RMSE         = {np.sqrt(np.mean((alpha_hat - true_alpha)**2)):.3f}")
```

## R

```{r}
#| eval: false
pkgload::load_all("../forestriesz/r/forestriesz")
use_python_forestriesz(file.path(getwd(), "../.venv/bin/python"))

set.seed(0)
n  <- 1500
x  <- runif(n)
pi <- 1 / (1 + exp(-(-0.02 * x - x^2 + 4 * log(x + 0.3) + 1.5)))
a  <- rbinom(n, 1, pi)
df <- data.frame(a = as.numeric(a), x = x)

# ATE / ATT use a multi-basis sieve, which is Python-only in v1.
# Call into Python directly via reticulate.
fr_py <- reticulate::import("forestriesz", convert = FALSE)
pd_py <- reticulate::import("pandas", convert = FALSE)
py_df <- pd_py$DataFrame(reticulate::r_to_py(list(a = as.numeric(a), x = x)))

fr <- fr_py$ForestRieszRegressor(
  estimand = fr_py$ATE(treatment = "a", covariates = list("x")),
  n_estimators = 500L, min_samples_leaf = 10L, random_state = 0L
)
fr$fit(py_df)
alpha_hat <- as.numeric(reticulate::py_to_r(fr$predict(py_df)))
```

:::

## Quickstart — AdditiveShift with the augmentation-style backend

For estimands without a canonical sieve, `AugForestRieszRegressor` is the right choice — no extra configuration.

```{python}
from forestriesz import AugForestRieszRegressor, AdditiveShift

# Continuous-treatment data
a_cont = rng.normal(0.5*x, 1.0)
df_cont = pd.DataFrame({"a": a_cont, "x": x})

afr = AugForestRieszRegressor(
    estimand=AdditiveShift(delta=0.5),
    n_estimators=500,
    min_samples_leaf=10,
    random_state=0,
)
afr.fit(df_cont)

alpha_hat = afr.predict(df_cont)
print(f"alpha_hat range: [{alpha_hat.min():.3f}, {alpha_hat.max():.3f}]")
```

The same estimator works on every built-in estimand at near-identical speed and quality vs the moment-style backend; trade off CIs for full agnosticism.

## Quickstart — TSM with confidence intervals

`TSM` uses a single-basis sieve, so v1's `predict_interval` works.

::: {.panel-tabset group="lang"}

## Python

```{python}
from forestriesz import ForestRieszRegressor, TSM

fr = ForestRieszRegressor(
    estimand=TSM(level=1, treatment="a", covariates=("x",)),
    n_estimators=500,
    honest=True,
    inference=True,
    random_state=0,
)
fr.fit(df)

alpha_hat = fr.predict(df)
lb, ub    = fr.predict_interval(df, alpha=0.05)
print(f"mean 95% CI width = {(ub - lb).mean():.3f}")
print(f"mean coverage of α̂ in [lb, ub] = {((lb <= alpha_hat) & (alpha_hat <= ub)).mean():.2f}")
```

## R

```{r}
#| eval: false
pkgload::load_all("../forestriesz/r/forestriesz")
use_python_forestriesz(file.path(getwd(), "../.venv/bin/python"))

fr <- ForestRieszRegressor$new(
  estimand = TSM(level = 1L, treatment = "a", covariates = "x"),
  n_estimators = 500L,
  honest = TRUE,
  inference = TRUE,
  random_state = 0L
)
fr$fit(df)
alpha_hat <- fr$predict(df)
ci <- fr$predict_interval(df, alpha = 0.05)
```

:::

## Hyperparameters

Forwarded to EconML's `BaseGRF`:

| Knob | Default | Notes |
|---|---|---|
| `n_estimators` | 100 | Number of trees. With `inference=True` must be divisible by `subforest_size` (default 4). |
| `max_depth` | None | Grow until leaves saturate `min_samples_leaf`. |
| `min_samples_leaf` | 5 | Tighter leaves give lower bias, higher variance. Paper guidance: 5–20. |
| `min_samples_split` | 10 | |
| `max_samples` | 0.45 | Per-tree subsample fraction. GRF default. |
| `honest` | False | Enable for valid-coverage CIs (50/50 sample split). |
| `inference` | False | Enable for `predict_interval`; preserves subforest structure. |
| `subforest_size` | 4 | Trees per subforest when computing inference variance. |
| `l2` | 0.01 | Ridge added to the per-leaf Jacobian. |
| `n_jobs` | -1 | Parallel tree training. |

## Save / load

```{python}
import tempfile, os
with tempfile.TemporaryDirectory() as tmp:
    save_dir = os.path.join(tmp, "fitted")
    fr.save(save_dir)

    loaded = ForestRieszRegressor.load(save_dir)
    pred_loaded = loaded.predict(df)
    print(f"max abs diff after round-trip: {np.max(np.abs(pred_loaded - alpha_hat)):.2e}")
```

For built-in estimands the sieve is auto-resolved on load. For custom sieves, repass the callables: `ForestRieszRegressor.load(path, riesz_feature_fns=my_basis)`.

## Sklearn integration

`ForestRieszRegressor` is a `BaseEstimator`, so it composes with `clone`, `GridSearchCV`, `cross_val_predict`, `Pipeline`. See [Tuning and Cross-Fitting](../tuning-and-cross-fitting.qmd).

## When the constant basis raises

For built-in estimands the per-row moment $m(W; 1)$ doesn't depend on $W$, so the constant basis is degenerate (every leaf would predict the same constant). The backend detects this and raises with a hint to use `riesz_feature_fns="auto"` (the default) or supply a custom sieve. The default does the right thing; explicit `riesz_feature_fns=None` is what raises.
