---
title: "Estimands"
---

{{< include _setup.qmd >}}

An **estimand** is a scalar summary of the data-generating distribution that you want to estimate. The estimands `rieszreg` handles all share the form
$$
\psi = \mathbb{E}\!\left[\,m(\mu)(Z, Y)\,\right]
$$
where $\mu(z) = \mathbb{E}[Y \mid Z = z]$ is the outcome regression and $m$ is an *operator* — a rule that maps the function $\mu$ to another function $m(\mu)$ that can also depend on the row $(Z, Y)$. For the average treatment effect (ATE) with binary treatment $A$, $m(\mu)(Z, Y) = \mu(1, X) - \mu(0, X)$ — `m` ignores $Y$ and produces a contrast of two evaluations of $\mu$.

An estimand object packages two things: the column names $\alpha$ is indexed by (`feature_keys`) and the operator itself (`m`). `m` is curried: it takes a candidate $\alpha$ and returns a function of one row $(z, y)$. The five built-in factories (`ATE`, `ATT`, `TSM`, `AdditiveShift`, `LocalShift`) cover most causal-inference targets; section [Custom estimands](#custom-estimands) shows how to write your own.

The factories live in `rieszreg`; `rieszboost`, `krrr`, `forestriesz`, and `riesznet` all re-export them so `from rieszboost import ATE`, `from krrr import ATE`, `from forestriesz import ATE`, and `from riesznet import ATE` resolve to the same object.

## Quickstart: ATE on the Lalonde NSW data

The NSW (National Supported Work) demonstration is the canonical Lalonde dataset: a job-training experiment for disadvantaged workers, plus a CPS observational comparison group. The goal here is the average treatment effect on 1978 earnings.

The columns we need:

- `treat`: binary treatment indicator (1 = NSW trainee, 0 = CPS comparison).
- `re78`: 1978 real earnings (the outcome $Y$).
- `age`, `educ`, `black`, `hisp`, `marr`, `nodegree`, `re74`, `re75`: pre-treatment covariates.

The estimand needs `treat` and the eight covariates. The outcome `re78` becomes `y` — a separate vector — at fit time.

::: {.panel-tabset group="lang"}

## Python

```{python}
import numpy as np, pandas as pd
from causaldata import nsw_mixtape, cps_mixtape
from rieszreg import ATE
from rieszboost import RieszBooster

COVARIATES = ["age", "educ", "black", "hisp", "marr", "nodegree", "re74", "re75"]

treated  = nsw_mixtape.load_pandas().data
treated  = treated[treated["treat"] == 1].copy()
controls = cps_mixtape.load_pandas().data
df = pd.concat([treated, controls], ignore_index=True)
df["treat"] = df["treat"].astype(float)

Z = df[["treat"] + COVARIATES]
y = df["re78"].astype(float)
print(f"n = {len(df)}  ({(df['treat']==1).sum()} treated, {(df['treat']==0).sum()} control)")
```

The estimand says exactly which dataframe columns feed $\alpha$: the treatment column and the eight covariates, in the order they appear in `feature_keys`.

```{python}
estimand = ATE(treatment="treat", covariates=tuple(COVARIATES))
estimand.feature_keys
```

Fit the Riesz representer. `y` is passed because the sklearn convention says so; the ATE operator ignores it.

```{python}
booster = RieszBooster(
    estimand=estimand,
    n_estimators=300, learning_rate=0.05, max_depth=4, reg_lambda=10.0,
    random_state=0,
).fit(Z, y)
```

Predict $\hat\alpha$ on the same rows and look at extreme-weight diagnostics.

```{python}
alpha_hat = booster.predict(Z)
print(f"min α̂ = {alpha_hat.min():.2f}, max α̂ = {alpha_hat.max():.2f}")
booster.diagnose(Z).summary()
```

To turn $\hat\alpha$ and a separately fit outcome regression $\hat\mu$ into a one-step or DML estimate of $\psi$, see the [DML / TMLE recipe](estimation/custom.qmd).

## R

```{r}
library(reticulate)
cd_py <- import("causaldata")
COVARIATES <- c("age", "educ", "black", "hisp", "marr", "nodegree", "re74", "re75")

treated  <- py_to_r(cd_py$nsw_mixtape$load_pandas()$data)
treated  <- treated[treated$treat == 1, ]
controls <- py_to_r(cd_py$cps_mixtape$load_pandas()$data)
df <- rbind(treated, controls)
df$treat <- as.numeric(df$treat)

Z <- df[, c("treat", COVARIATES)]
y <- df$re78
cat(sprintf("n = %d  (%d treated, %d control)\n",
            nrow(df), sum(df$treat == 1), sum(df$treat == 0)))
```

```{r}
booster <- RieszBooster$new(
  estimand = ATE(treatment = "treat", covariates = COVARIATES),
  n_estimators = 300L, learning_rate = 0.05, max_depth = 4L, reg_lambda = 10.0,
  random_state = 0L
)
booster$fit(Z, y)
```

```{r}
alpha_hat <- booster$predict(Z)
cat(sprintf("min alpha = %.2f, max alpha = %.2f\n",
            min(alpha_hat), max(alpha_hat)))
booster$diagnose(Z)$summary
```

:::

## Built-in estimands

| Estimand | $m(\mu)(z, y)$ | Notes |
|---|---|---|
| `ATE(treatment, covariates)` | $\mu(1, x) - \mu(0, x)$ | Average treatment effect — see [`rieszboost/examples/lee_schuler/binary_dgp.py`](https://github.com/rieszreg/rieszboost/blob/main/examples/lee_schuler/binary_dgp.py). |
| `ATT(treatment, covariates)` | $a \cdot (\mu(1, x) - \mu(0, x))$ | ATT *partial-estimand* surface. Full ATT divides by $P(A=1)$ — combine with a delta-method EIF downstream. |
| `TSM(level, treatment, covariates)` | $\mu(\text{level}, x)$ | Treatment-specific mean — see [`rieszboost/examples/tsm.py`](https://github.com/rieszreg/rieszboost/blob/main/examples/tsm.py). |
| `AdditiveShift(delta, ...)` | $\mu(a + \delta, x) - \mu(a, x)$ | Continuous-treatment shift effect — see [`rieszboost/examples/lee_schuler/continuous_dgp.py`](https://github.com/rieszreg/rieszboost/blob/main/examples/lee_schuler/continuous_dgp.py). |
| `LocalShift(delta, threshold, ...)` | $\mathbf{1}\{a < \text{threshold}\} \cdot (\mu(a + \delta, x) - \mu(a, x))$ | LASE *partial-estimand* surface; same caveat as ATT. |

The R and Python factories take the same arguments.

::: {.panel-tabset group="lang"}

## Python

```{python}
from rieszreg import ATE, ATT, TSM, AdditiveShift, LocalShift
ATE(treatment="a", covariates=("x",)).name
TSM(level=1, treatment="a", covariates=("x",)).name
LocalShift(delta=0.3, threshold=0.7).name
```

## R

```{r}
ATE(treatment = "a", covariates = "x")
TSM(level = 1, treatment = "a", covariates = "x")
LocalShift(delta = 0.3, threshold = 0.7)
```

:::

## Mapping your dataframe to an estimand

An estimand object carries two pieces: the column names $\alpha$ is indexed by (`feature_keys`) and the operator `m(alpha)(z, y)`.

`feature_keys` declares the input schema of $\alpha$ — the columns the underlying learner consumes when scoring a row. The orchestrator slices these columns out of `Z` per row to build the feature matrix the learner sees.

The outcome `Y` is **not** part of `feature_keys`. It flows in the sklearn-style way: a separate vector passed as the second argument to `.fit(Z, y)`. The orchestrator threads it through to `m(alpha)(z, y)`. Built-in estimands ignore `y`; pass it anyway when following sklearn convention.

For the Lalonde quickstart above, the wiring looks like this:

| Concept | Lalonde column | Where it goes |
|---|---|---|
| Treatment $A$ | `treat` | `ATE(treatment="treat", ...)` — a `feature_key`. |
| Covariates $X$ | `age`, `educ`, `black`, `hisp`, `marr`, `nodegree`, `re74`, `re75` | `ATE(covariates=(...))` — also `feature_keys`. |
| Outcome $Y$ | `re78` | `.fit(Z, y=df["re78"])`. |

The `Z` dataframe must contain every name in `feature_keys`; extra columns are ignored. With an `ndarray`, columns are interpreted in `feature_keys` order.

## Theory: how the augmented dataset is derived

This section explains the math that lets `rieszreg` evaluate the Riesz loss without ever materializing $\alpha_0$. Skip on a first read.

While the Riesz representer theorem applies to any continuous linear functional, the estimands `rieszreg` supports are those whose $m$ reduces to a finite linear combination of point evaluations of $\mu$. For example, the ATE is $\mathbb{E}[\mu(1, X) - \mu(0, X)]$, which reduces to two point evaluations $\mu(1, x)$ and $\mu(0, x)$ for each row's covariates $x$. We call these **finite-evaluation estimands**.

Formally, an estimand $\psi = \mathbb{E}[m(\mu)(Z, Y)]$ is finite-evaluation if there exist fixed functions $a: \mathcal Z \times \mathcal Y \to \mathcal Z^k$ (the **point generator**) and $c: \mathcal Z \times \mathcal Y \to \mathbb R^k$ (the **weight function**) such that
$$
m(\mu)(z, y) = c(z, y)^\top \vec\mu(a(z, y)),
$$
where $\vec\mu$ denotes elementwise application of $\mu$ to the vector-valued input. For ATE, $k = 2$, $a(z, y) = ((1, z_x), (0, z_x))$, and $c(z, y) = (1, -1)$.

For these estimands the empirical Bregman-Riesz loss simplifies. Define the $k$ pseudo-datapoints $Z_{ij} = a_j(Z_i, Y_i)$ with coefficients $C_{ij} = c_j(Z_i, Y_i)$ for each row $i$. Then
$$
\begin{aligned}
\mathcal L_{h,n}(\alpha)
&= \sum_{i=1}^n \left[\tilde h(\alpha(Z_i)) - m(h' \circ \alpha)(Z_i, Y_i)\right] \\
&= \sum_{i=1}^n \left[\tilde h(\alpha(Z_i)) - \sum_{j=1}^k C_{ij} \cdot h'(\alpha(Z_{ij}))\right].
\end{aligned}
$$

The loss depends on $\alpha$ only through its values at $\{Z_i\} \cup \{Z_{ij}\}$, at most $n(k+1)$ points. The coefficients $C_{ij}$ depend on data but not on $\alpha$, so they can be computed once as a preprocessing step.

### The augmented dataset

The cleanest way to keep track of the loss, derivatives, etc. during training is to materialize an **augmented dataset** of the points $\{Z_i\} \cup \{Z_{ij}\}$ along with the appropriate coefficients $C_{ij}$.

Some pseudo-datapoints coincide with the original observation. For ATE, $Z_{i1} = (1, x_i)$ and $Z_{i2} = (0, x_i)$. A treated unit ($a_i = 1$) has $Z_{i1} = Z_i$, so two of the three evaluation points are the same. Let $D_{ij} = \mathbf{1}[Z_{ij} = Z_i]$ indicate overlap. Splitting the $h'$ sum into overlapping and non-overlapping terms:
$$
\mathcal L_{h,n}(\alpha) = \sum_{i=1}^n \left[\tilde h(\alpha(Z_i)) - \sum_{j:\, D_{ij}=1} C_{ij}\, h'(\alpha(Z_i))\right] - \sum_{i=1}^n \sum_{j:\, D_{ij}=0} C_{ij}\, h'(\alpha(Z_{ij})).
$$

At each distinct evaluation point, the loss combines a $\tilde h$-weight and an $h'$-weight into one term. `Estimand.augment` collects these distinct points for each row and stores their coefficients. Let $r$ index an enumeration of the augmented data points. Each augmented row $r$ corresponds to either a $Z_{ij}$ or a $Z_i$ for some original row $i$; write $i_r$ for that origin row. The two weights are
$$
D_r = \mathbf{1}[Z_r = Z_{i_r}],
\qquad
C_r = -\sum_{j \,:\, Z_{i_r, j} = Z_r} C_{i_r, j}.
$$

The $\tilde h$-weight $D_r$ is $1$ exactly when $Z_r$ is the original observation; the $h'$-weight $C_r$ collects $-C_{ij}$ from every pseudo-datapoint of row $i_r$ that lands at $Z_r$ — an empty sum (giving $C_r = 0$) when no pseudo-datapoint coincides with $Z_{i_r}$. The per-row contribution to the empirical loss is $L_r = D_r\, \tilde h(\alpha(Z_r)) + C_r\, h'(\alpha(Z_r))$, and $\mathcal L_{h,n}(\alpha) = \sum_r L_r$.

For ATE with an untreated unit ($a_i = 0$): before merging there are three evaluation points — $Z_i = (0, x_i)$ and pseudo-datapoints $(1, x_i)$, $(0, x_i)$. The pseudo-datapoint $(0, x_i)$ coincides with $Z_i$ ($D_{i2} = 1$), so they merge into one entry with $\tilde h$-weight $1$ and $h'$-weight $-C_{i2} = +1$. The other pseudo-datapoint $(1, x_i)$ stays separate with $\tilde h$-weight $0$ and $h'$-weight $-C_{i1} = -1$. The result is two augmented rows instead of three. A treated unit merges the other pseudo-datapoint instead, also producing two rows.

It can also happen that none of the pseudo-datapoints coincide with $Z_i$ — the union $\{Z_i\} \cup \{Z_{ij}\}$ then has $k+1$ distinct entries and no merging occurs. For example, `TSM(level=1)` on an untreated row ($a_i = 0$) has $Z_i = (0, x_i)$ and a single pseudo-datapoint $(1, x_i)$; they are distinct, so $Z_i$ enters the augmented dataset as its own row with $\tilde h$-weight $1$ and $h'$-weight $0$ (no pseudo-datapoint lands there), and $(1, x_i)$ enters as a separate row with $\tilde h$-weight $0$ and $h'$-weight $-C_{i1} = -1$. The general rule is the same in every case: include $Z_i$ once, then add each pseudo-datapoint $Z_{ij}$ — merging into an existing entry when $Z_{ij}$ matches a point already in the dataset for row $i$, otherwise as a new row.

To make this concrete, take the ATE estimand on $n=3$ rows:

| $i$ | $A_i$ | $X_i$ |
|----:|------:|------:|
| 1   | 0     | 0.2   |
| 2   | 1     | 0.5   |
| 3   | 1     | 0.8   |

For each row, the three evaluation points $\{Z_i,\, Z_{i1} = (1, X_i),\, Z_{i2} = (0, X_i)\}$ collapse to two distinct entries after merging. The augmented dataset has $2n = 6$ rows; each row $r$ carries the evaluation point $(A_r, X_r)$, the $\tilde h$-weight $D_r$, the $h'$-weight $C_r$, and contributes a term $L_r = D_r \cdot \tilde h(\alpha(A_r, X_r)) + C_r \cdot h'(\alpha(A_r, X_r))$ to the empirical loss:

| from $i$ | $A_r$ | $X_r$ | $D_r$ | $C_r$ | $L_r$ |
|---------:|------:|------:|------:|------:|:------|
| 1        | 0     | 0.2   | $1$   | $+1$  | $\tilde h(\alpha(0, 0.2)) + h'(\alpha(0, 0.2))$ |
| 1        | 1     | 0.2   | $0$   | $-1$  | $-h'(\alpha(1, 0.2))$ |
| 2        | 1     | 0.5   | $1$   | $-1$  | $\tilde h(\alpha(1, 0.5)) - h'(\alpha(1, 0.5))$ |
| 2        | 0     | 0.5   | $0$   | $+1$  | $+h'(\alpha(0, 0.5))$ |
| 3        | 1     | 0.8   | $1$   | $-1$  | $\tilde h(\alpha(1, 0.8)) - h'(\alpha(1, 0.8))$ |
| 3        | 0     | 0.8   | $0$   | $+1$  | $+h'(\alpha(0, 0.8))$ |

Rows $i=2$ and $i=3$ are treated, so the merging happens on the $(1, X_i)$ pseudo-datapoint instead of $(0, X_i)$, flipping which row carries the $\tilde h$-weight. Summing the $L_r$ column recovers the empirical loss:
$$
\mathcal L_{h,n}(\alpha) = \sum_{r=1}^{6} L_r = \sum_{i=1}^{3} \big[\tilde h(\alpha(Z_i)) - h'(\alpha(1, X_i)) + h'(\alpha(0, X_i))\big].
$$

`Estimand.augment` produces this table directly: it stores the two coefficients as `is_original` (the $\tilde h$-weight $D_r$) and `potential_deriv_coef` (the $h'$-weight $C_r$) on the returned `AugmentedDataset`.

```{python}
import numpy as np
import pandas as pd
from rieszreg import ATE

df = pd.DataFrame({"a": [0, 1, 1], "x": [0.2, 0.5, 0.8]})
df
```

```{python}
estimand = ATE(treatment="a", covariates=("x",))
aug = estimand.augment(df[["a", "x"]].to_numpy(dtype=float))

aug_df = pd.DataFrame({
    "i": aug.origin_index + 1,
    "A": aug.features[:, 0],
    "X": aug.features[:, 1],
    "D": aug.is_original,
    "C": aug.potential_deriv_coef,
})
aug_df
```

Pick a candidate $\alpha(a, x) = a + x$ and the squared loss ($\tilde h(t) = t^2$, $h'(t) = 2t$). The per-row contribution becomes $L_r = D \cdot \alpha^2 + 2 C \cdot \alpha$:

```{python}
F = aug_df["A"] + aug_df["X"]                    # α evaluated at each augmented point
aug_df["L_r"] = aug.is_original * F**2 + 2.0 * aug.potential_deriv_coef * F
aug_df
```

```{python}
aug_df["L_r"].sum()                              # empirical loss
```

The same number drops out of the original row-wise form $\sum_i [\alpha(Z_i)^2 - 2\alpha(1, X_i) + 2\alpha(0, X_i)]$:

```{python}
((df["a"] + df["x"])**2 - 2*(1 + df["x"]) + 2*(0 + df["x"])).sum()
```

## Custom estimands

To fit a finite-evaluation functional not covered by the built-ins, write the operator `m(alpha)(z, y)` and wrap it in a `FiniteEvalEstimand`. The tracer extracts the linear-form structure automatically.

`m` must be a finite linear combination of point evaluations of $\alpha$, built with `+`, `-`, scalar `*`, and calls to `alpha(...)`. Squaring, adding constants, integrals, and other non-linear operations raise.

::: {.callout-note title="Python-only"}
Custom `m()` is Python-only. From R, define `m` in Python via `reticulate::py_run_string` and pass the resulting `FiniteEvalEstimand` to your estimator. See the [R interface page](r-interface.qmd) for the recipe.
:::

```{python}
from rieszreg import FiniteEvalEstimand
from rieszboost import RieszBooster
import numpy as np, pandas as pd

# A weighted contrast: 2x the ATE at level a=2 minus the level at a=1.
def m_weighted_contrast(alpha):
    def inner(z, y=None):
        return 2.0 * (alpha(a=2, x=z["x"]) - alpha(a=0, x=z["x"])) \
               - alpha(a=1, x=z["x"])
    return inner

est = FiniteEvalEstimand(
    feature_keys=("a", "x"),
    m=m_weighted_contrast,
    name="WeightedContrast",
)
print(est)
```

The inner function takes `(z, y)` even when `y` is unused — that signature is the contract every custom `m` follows. Pass the estimand to any backend's estimator class:

```{python}
rng = np.random.default_rng(0)
df = pd.DataFrame({
    "a": rng.choice([0, 1, 2], size=500).astype(float),
    "x": rng.uniform(0, 1, 500),
})

booster = RieszBooster(estimand=est, n_estimators=100, max_depth=3).fit(df)
booster.predict(df)[:5]
```

### What the tracer rejects

The errors point at the offending operation:

```{python}
#| error: true
from rieszreg import FiniteEvalEstimand, trace

def bad_squared(alpha):
    return lambda z, y=None: alpha(a=1, x=z["x"]) ** 2  # nonlinear in alpha

trace(FiniteEvalEstimand(feature_keys=("a", "x"), m=bad_squared), {"a": 0, "x": 0.5})
```

```{python}
#| error: true
def bad_constant(alpha):
    return lambda z, y=None: alpha(a=1, x=z["x"]) + 1.0  # constant offset

trace(FiniteEvalEstimand(feature_keys=("a", "x"), m=bad_constant), {"a": 0, "x": 0.5})
```

For estimands with integrals or derivatives of $\alpha$ (outside the finite-linear-form algebra), Monte-Carlo-sample and approximate as a finite sum.

## Initialization

Every `RieszEstimator` subclass takes an `init=` constructor arg that controls the α-space starting value of the model. The orchestrator converts `init` to an η-space `base_score` via `loss.alpha_to_eta`; backends use `base_score` as a prediction offset.

The default — `init=None` — sets α to the **empirical loss-minimizing constant**. For any Bregman loss with strictly convex $h$, the constant $a^*$ that minimizes the per-row Riesz loss
$$
P_n L(a) \;=\; \tilde h(a) \;-\; h'(a) \cdot \bar m, \qquad \bar m = P_n[m(\alpha = 1)(Z, Y)]
$$
satisfies the FOC $\tilde h'(a) = h''(a) \bar m$. Using $\tilde h(t) = t h'(t) - h(t)$ we get $\tilde h'(t) = t h''(t)$, so the FOC collapses to $a^* = \bar m$. The result is **loss-independent** — the only loss-specific step is projecting $\bar m$ into the loss's α-domain when its codomain is restricted. Each `Loss` exposes `best_constant_init(m_bar)` for that projection.

For the built-in estimands $\bar m$ is structurally constant (doesn't depend on the data realization):

| Estimand | $m(\alpha = 1)(Z, Y)$ per row | $\bar m$ |
|---|---|---|
| `ATE`, `ATT`, `AdditiveShift`, `LocalShift` | $1 + (-1) = 0$ | $0$ |
| `TSM(v)` | $1$ | $1$ |

For custom estimands $\bar m$ can depend on the data; the orchestrator computes it from the training rows after the validation split (no peeking) by tracing each row and averaging the sum of trace coefficients.

Override the default by passing a float:

```python
booster = RieszBooster(estimand=ATE(), init=0.0)        # explicit zero baseline
booster = RieszBooster(estimand=TSM(level=1), init=2.5) # informed prior
```

::: {.callout-note title="Boundary cases"}
For `BernoulliLoss` + `TSM` (or any density-ratio estimand with $\bar m = 1$) the loss-minimizing constant sits on the upper boundary, and the projection clips to $1 - \varepsilon$. The sigmoid is saturated at that init and gradients are tiny, so iterative training may stall. Pass `init=0.5` (or another interior value) when this matters. The `KLLoss` / `BoundedSquaredLoss` boundaries can hit the same issue when $\bar m$ lands exactly on a bound.
:::

## Estimand subclass taxonomy and future work

Today every estimand `rieszreg` handles is a `FiniteEvalEstimand`: $m$ reduces to a finite linear combination of point evaluations of $\mu$. The taxonomy of $m$-operators outside this class — integrals, derivatives, and others — is still being designed. Likely subclasses:

- **Integrals.** $m(\mu)(z, y) = \int \mu(a', x)\,h(a' \mid z)\,da'$ for a known density $h$. The classic example is a stochastic-intervention policy. Approximate via Monte Carlo to reduce to the finite-evaluation case (with $K$ samples per row), or support natively via a backend that integrates $\alpha$ analytically.
- **Derivatives.** $m(\mu)(z, y) = \partial_a \mu(a, x)\big|_{a = z_a}$. Finite-difference approximations reduce to the finite-evaluation case; backends with autodiff can support directly.
- **Other linear operators on $\mu$.** Any $m$ that satisfies linearity on $\mu$ (additivity and scalar homogeneity) but doesn't fit the patterns above.

A previously released `StochasticIntervention` factory is currently stubbed (raises `NotImplementedError`). A reintroduction will land once the integral-vs-finite-evaluation distinction is wired into the type hierarchy.
